Innovation and elaboration on the avian tree of life - supplementary materials

### 2023-06-01 Manuscript/main.tex
Update main.tex

Various approaches have been developed to measure properties of trait matrices \cite{hansen2008measuring, pavlicev2009measuring, watanabe2022}, broadly focused on describing the size, shape, and orientation of trait divergence \cite{phillips1999}.
Our specific focus is on divergence of species and clades along or away from the phylogenetic line of least resistance.
We adopt concepts developed in the specific context of signal evolution to describe these alternative routes to divergence \cite{endler2005animal}.
Specifically, Endler \textit{et al}. (2005) refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, is expected to be a major driver of phenotypic evolution.
In contrast, Endler \textit{et al}. (2005) refer to evolution away from the major axis of phenotypic variation, either in random directions or along limited axes, as \textit{innovation} \cite{endler2005animal}.
Relative to elaboration, innovation is more likely to generate phenotypic novelty, but, if directions of phenotypic evolution are constrained, would not be expected to be a primary driver of phenotypic diversity among closely related species.
We apply these terms to i) species divergence within clades, and ii) clade divergence within larger clades (Figure 1).
Both innovation and elaboration could in principle occur at the macro- or mega-evolutionary scale.
We expect that elaboration is more important at macroevolutionary scales (species diverging within clades) and innovation more important at megaevolutionary scale (divergence among clades).
For our bird beak data, macroevolutionary scales refer to the beak shapes of species elaborating or innovating relative to the phylogenetic line of least resistance of their taxonomic order, and megaevolutionary scales refer to avian orders elaborating or innovating relative to the phylogenetic line of least resistance for the entire class Aves.
For clarity throughout the manuscript, we refer to elaboration and innovation as the general concept described above but use the following terms to distinguish macroevolutionary scale (species) and megaevolutionary scale (clade) elaboration and innovation: $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ and $innovation_{clade}$.
We describe the algebraic formulation of elaboration and innovation for species and clades in detail in the Methods section.
Using these novel statistical tools, we estimate the relative contribution of elaboration and innovation to the remarkable global radiation and diversity of bird beaks.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/correlations_fable_orders.pdf}
\caption{.}
\label{fable_correlations}
\end{figure}
\bigskip

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/ellipses.pdf}
\caption{.}
\label{Fig:ellipses}
\end{figure}
\bigskip

>>>>

Various approaches have been developed to measure properties of trait matrices \cite{hansen2008measuring, pavlicev2009measuring, watanabe2022}, broadly focused on describing the size, shape, and orientation of trait divergence \cite{phillips1999}.
Our specific focus is on divergence of species and clades along or away from the phylogenetic line of least resistance.
We adopt concepts developed in the specific context of signal evolution to describe these alternative routes to divergence \cite{endler2005animal}.
Specifically, Endler \textit{et al}. (2005) refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, is expected to be a major driver of phenotypic evolution.
In contrast, Endler \textit{et al}. (2005) refer to evolution away from the major axis of phenotypic variation, either in random directions or along limited axes, as \textit{innovation} \cite{endler2005animal}.
Relative to elaboration, innovation is more likely to generate phenotypic novelty, but, if directions of phenotypic evolution are constrained, would not be expected to be a primary driver of phenotypic diversity among closely related species.
We apply these terms to i) species divergence within clades, and ii) clade divergence within larger clades (Fig. \ref{Fig:cheatsheet}).
Both innovation and elaboration could in principle occur at the macro- or mega-evolutionary scale.
We expect that elaboration is more important at macroevolutionary scales (species diverging within clades) and innovation more important at megaevolutionary scale (divergence among clades).
For our bird beak data, macroevolutionary scales refer to the beak shapes of species elaborating or innovating relative to the phylogenetic line of least resistance of their taxonomic order, and megaevolutionary scales refer to avian orders elaborating or innovating relative to the phylogenetic line of least resistance for the entire class Aves.
For clarity throughout the manuscript, we refer to elaboration and innovation as the general concept described above but use the following terms to distinguish macroevolutionary scale (species) and megaevolutionary scale (clade) elaboration and innovation: $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ and $innovation_{clade}$.
We describe the algebraic formulation of elaboration and innovation for species and clades in detail in the Methods section.
Using these novel statistical tools, we estimate the relative contribution of elaboration and innovation to the remarkable global radiation and diversity of bird beaks.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/correlations_fable_orders.pdf}
\caption{.}
\label{fable_correlations}
\end{figure}
\bigskip

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/figure_phylo_spiro_dark.pdf}
\caption{.}
\label{Fig:ellipses}
\end{figure}
\bigskip
